### 2017-06-07 src/5_2_Proposal.tex
change title 5-2 and date
chapter 5 src/5_2 Icpram 2015 ready

\chapter{Online VECM proposal}
Financial time series are known for their non-stationary behaviour. However,
sometimes they exhibit some stationary linear combinations. When this happens,
it is said that those time series are cointegrated. The Vector Error Correction
Model (VECM) is an econometric model which characterises the joint dynamic
behaviour of a set of cointegrated variables in terms of forces pulling towards
equilibrium. In this study, we propose an Online VEC model (OVECM) which
optimises how model parameters are obtained using a sliding window of the most
recent data. Our proposal also takes advantage of the long-run relationship
between the time series in order to obtain improved execution times. Our
proposed method is tested using four foreign exchange rates with a frequency of
1-minute, all related to the USD currency base. OVECM is compared with VECM and
ARIMA models in terms of forecasting accuracy and execution times. We show that
OVECM outperforms ARIMA forecasting and enables execution time to
be reduced considerably while maintaining good accuracy levels compared with
VECM.

@@

\section{The problem}
In finance, it is common to find variables with a long-run and/or a short-run
equilibrium relationship. This relationship is called cointegration and it
reflects the idea of that some set of variables cannot wander too far away from
each other.

Cointegration means that one or more combinations of these variables is
stationary even though individually they are not.

Some financial models, such as  vector error correction model (VECM), take advantage of
this property and describe the joint behaviour of several cointegrated
financial instruments.

VECM is a special case of the vector autoregressive model (VAR). VAR is a linear
model which express future values in terms of its own history and other
variables past values. If cointegration exists, VECM allows to introduce an
error correction term due to its cointegration estimation error.

VECM as well as VAR model parameters are obtained using ordinary least squares
method (OLS). Since OLS involves many calculations, the parameter estimation
method is computationally expensive when the number of past values and
observations considered increases. Moreover, OLS is an ill-posed problem which
admits an infinite number of solutions.

Ridge regression method (RR) tackles this ill-posed problem and is usually
formulated instead of OLS. RR includes a regularisation parameter that leads to
better generalisation capability. However, RR is still computationally
expensive dealing with large datasets. Recently, online learning algorithms
have been proposed to solve problems with large data sets because of their
simplicity and their ability of updating the model when new data is available.

We propose an online formulation of the VEC model based on the
AAR method considering only a sliding window of the most recent data. The
algorithm introduces matrix optimisations in order to reduce the number of
operations. Our method is later tested with financial data from the foreign
exchange market.

\section{Methodology} \label{sec:methodology}

@@

Algorithm~\ref{alg:proposal} shows our OVECM proposal which considers the
following:

\begin{itemize}
\item The function \texttt{getJohansen} returns cointegration vectors given by
the Johansen method considering the trace statistic test at 95\%
significance level.
\item The function \texttt{vecMatrix} returns the matrices~(\ref{eq:vecA})
and~(\ref{eq:vecY}) that allows VECM to be solved.
\item The function \texttt{vecMatrixOnline} returns the
matrices~(\ref{eq:vecA}) and~(\ref{eq:vecY}) aggregating new data and removing
the old one, avoiding calculation of the matrix $\mathbf{A}$ from scratch.
\item Out-of-sample outputs are saved in the variables 
$\mathbf{Y}_{\text{true}}$ and $\mathbf{Y}_{\text{pred}}$.
\item The model is solved using OLS.
\item In-sample outputs are saved in the variables $\Delta
\mathbf{y}_{\text{true}}$ and $\Delta \mathbf{y}_{\text{pred}}$.
\item The function \texttt{mape} gets the in-sample MAPE for the $l$ time
series.
\item Cointegration vectors are obtained at the beginning and when they are required to be updated. This updating is decided based on the in-sample MAPE of the last $n$ inputs. The average of all
MAPEs are stored in the variable $e$. If the average of MAPEs
($\texttt{mean}(e)$) is above a certain error given by the mean\_error threshold, cointegration vectors are updated.
\item If new cointegration vectors are required, the function
\texttt{vecMatrixUpdate} only updates the corresponding columns of matrix
$\mathbf{A}$ affected by those vectors (see equation~\ref{eq:vecA}).
\end{itemize}

\begin{algorithm}[ht]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
$p$: number of past values \\
$L$: sliding window size ($L<N$) \\
$\text{mean\_error}$: MAPE threshold \\
$n$: interpolation points to obtain MAPE \\
\ENSURE  $\,$ \\
$\{\mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
\FOR { $i =0$ to $N-L$ }
    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L]$
	\IF {i = 0 }
	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
	    \STATE{$[\mathbf{A} \quad \mathbf{Y}] \gets
        \texttt{vecMatrix}(\mathbf{y}_i,p,v)$}
	\ELSE
	    \STATE{$[\mathbf{A} \quad \mathbf{Y}] \gets
        \texttt{vecMatrixOnline}(\mathbf{y}_i,p,v,\mathbf{A},\mathbf{Y})$}
        \STATE $\Delta \mathbf{Y}_{\text{pred}}[i] \gets \mathbf{A}[-1,:] \times \mathbf{X}$
    \ENDIF
    \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{Y})$
    \STATE $e \gets \texttt{mape}(\mathbf{Y}[-n,:],\mathbf{A}[-n,:] \times \mathbf{X})$
    \IF {$\texttt{mean}(e) > \text{mean\_error}$}
	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
	    \STATE{$\mathbf{A} \gets
        \texttt{vecMatrixUpdate}(\mathbf{y}_i,p,v,\mathbf{A})$}
        \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{Y})$
    \ENDIF
\ENDFOR
\STATE $\mathbf{Y}_{\text{true}} \gets \mathbf{Y}[L+1:N]$
\STATE $\mathbf{Y}_{\text{pred}}\gets 
\mathbf{Y}[L:N-1] +\Delta \mathbf{Y}_{\text{pred}}$
\end{algorithmic}
\caption{OVECM: Online VECM}
\label{alg:proposal}
\end{algorithm}

A pseudocode of the algorithm \ref{alg:proposal} is detailed in the appendix \ref{fig:pseudocodigo}.

@@

\section{Results} \label{sec:results}

@@

\begin{table}[h!]
\caption{Unit roots tests}
\label{tab:adf}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
& \textbf{Statistic} & \textbf{Critical value} & \textbf{Result}\\
\hline
EURUSD          & -0.64 & -1.94 & True       \\
$\Delta$EURUSD & -70.45   & -1.94 & False       \\
GBPUSD          & -0.63   & -1.94 & True          \\
$\Delta$GBPUSD & -54.53   & -1.94 & False       \\
CHFUSD          & -0.88   & -1.94 & True         \\
$\Delta$CHFUSD & -98.98   & -1.94 & False       \\
JPYUSD          & -0.65 & -1.94 & True        \\
$\Delta$JPYUSD & -85.78 & -1.94 & False     \\ 
\hline
\end{tabular}
\end{center}
\end{table}

Table~\ref{tab:adf} shows that all currency rates cannot reject the unit root
test but they rejected it with their first differences. This means that all of
them are I(1) time series and we are allowed to use VECM and therefore OVECM.

@@

ARIMA parameters were also obtained using AIC.
Parameters optimisation is presented in table~\ref{tab:params}:

\begin{table}[ht]
\caption{Parameters optimisation. VECM order and ARIMA parameters were selected
using AIC.}
\label{tab:params}
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Windows size $L$ & VECM & ARIMA\\
 $L$ & order $(p)$ & order $(p,d,q)$ \\
\hline
100 & 2 & p=2,d=1,q=1\\
400 & 5 & p=1,d=1,q=1\\
700 & 3 &p=2,d=1,q=1\\
1000 &3 & p=2,d=1,q=1\\
\hline
\end{tabular}
\end{center}
\end{table}

@@

Table~\ref{tab:stats} shows in-sample and out-of-sample performance measures:
MAPE, MAE and RMSE for OVECM, SLVECM and SLARIMA. Test were done using the
parameters defined in table \ref{tab:params}.  We can see that OVECM has very
similar performance than SLVECM and this support the theory that cointegration
vectors vary little in time. Moreover, OVECM also out performed SLARIMA based on
these three performance measures.

@@

\section{Conclusions} \label{sec:conclusions}

>>>>

\chapter{An Online Vector Error Correction Model for Exchange Rates Forecasting}

The Vector Error Correction Model (VECM) is an econometric model which
characterises the joint dynamic behaviour of a set of cointegrated variables in
terms of forces pulling towards equilibrium. In the previous proposal, the
Adaptive VECM (AVECM) showed that to be able to use VECM at every time step can
be used for forecasting purposes. However, to avoid extensive calculations, a
parallel approach was implemented.

In this study, an Online VEC model (OVECM) is proposed, which optimises how
model parameters are obtained using a sliding window of the most recent data.
OVECM, unlile AVECM, updates the parameters at each step, instead of obtaining
new ones. This proposal also takes advantage of the long-run relationship
between the time series in order to obtain improved execution times.

Our proposed method is tested using four foreign exchange rates with a frequency
of 1-minute, all related to the USD currency base.

OVECM is compared with VECM and ARIMA models in terms of forecasting accuracy
and execution times. We show that OVECM outperforms ARIMA forecasting and
enables execution time to be reduced considerably while maintaining good
accuracy levels compared with VECM.

This work is published in \cite{icpram15}.

@@

\newpage
\section{The problem}

Both VECM and VAR model parameters are obtained using ordinary least squares
(OLS) method. Since OLS involves many calculations, the parameter estimation
method is computationally expensive when the number of past values and
observations increases. Moreover, obtaining cointegration vectors is also an
expensive routine.

Recently, online learning algorithms have been proposed to solve problems with
large data sets because of their simplicity and their ability to update the
model when new data is available. The study presented by ~\cite{arce+salinas2012}
applied this idea using ridge regression.

There are several popular online methods such as perceptron \cite{rosenblatt58},
passive-aggressive \cite{crammerETall2006}, stochastic gradient
descent \cite{zhang2004}, aggregating algorithm \cite{vovk2001} and the second
order perceptron \cite{cesa-bianchi2005}.  In \cite{cesa-bianchi2006}, an
in-deph analysis of online learning is provided.

In this proposal, an online formulation of the VECM called Online VECM (OVECM)
is presented. OVECM is a lighter version of VECM which considers only a sliding
window of the most recent data and introduces matrix optimizations in order to
reduce the number of operations and therefore execution times. OVECM also takes
into account the fact that cointegration vector space doesn't experience large
changes with small changes in the input data.

OVECM is later compared against VECM and ARIMA models using four currency rates
from the foreign exchange market with 1-minute frequency. VECM and ARIMA models
were used in an iterative way in order to allow fair comparison. Execution times
and forecast performance measures MAPE, MAE and RMSE were used to compare all
methods.

Model effectiveness is focused on out-of-sample forecast rather than in-sample
fitting. This criteria allows the OVECM prediction capability to be expressed
rather than just explaining data history.

The next sections are organized as follows: the OVECM algorithm proposed is
presented in section~\ref{sec:52methodology}. Section~\ref{sec:52results} gives a
description of the data used and the tests carried on to show accuracy and time
comparison of our proposal against the traditional VECM and
section~\ref{sec:52conclusions} includes conclusions and a proposal for future
study.

\section{Methodology} \label{sec:52methodology}

@@

Algorithm~\ref{alg:52proposal} shows our OVECM proposal.
\begin{algorithm}[!htp]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
$p$: number of past values \\
$L$: sliding window size ($L<N$) \\
$\text{mean\_error}$: MAPE threshold \\
$n$: interpolation points to obtain MAPE \\
\ENSURE  $\,$ \\
$\{\mathbf{y}_{\text{pred}}[L+1],\dots,\mathbf{y}_{\text{pred}}[N]\}$: model predictions 
\FOR { $i =0$ to $N-L$ }
    \STATE $\mathbf{y}_i \gets \mathbf{y}[i:i+L]$
	\IF {i = 0 }
	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
	    \STATE{$[\mathbf{A} \quad \mathbf{Y}] \gets
        \texttt{vecMatrix}(\mathbf{y}_i,p,v)$}
	\ELSE
	    \STATE{$[\mathbf{A} \quad \mathbf{Y}] \gets
        \texttt{vecMatrixOnline}(\mathbf{y}_i,p,v,\mathbf{A},\mathbf{Y})$}
        \STATE $\Delta \mathbf{Y}_{\text{pred}}[i] \gets \mathbf{A}[-1,:] \times \mathbf{X}$
    \ENDIF
    \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{Y})$
    \STATE $e \gets \texttt{mape}(\mathbf{Y}[-n,:],\mathbf{A}[-n,:] \times \mathbf{X})$
    \IF {$\texttt{mean}(e) > \text{mean\_error}$}
	    \STATE{$v \gets \texttt{getJohansen}(\mathbf{y}_i,p)$}
	    \STATE{$\mathbf{A} \gets
        \texttt{vecMatrixUpdate}(\mathbf{y}_i,p,v,\mathbf{A})$}
        \STATE $\mathbf{X} \gets \texttt{OLS} (\mathbf{A},\mathbf{Y})$
    \ENDIF
\ENDFOR
\STATE $\mathbf{Y}_{\text{true}} \gets \mathbf{Y}[L+1:N]$
\STATE $\mathbf{Y}_{\text{pred}}\gets 
\mathbf{Y}[L:N-1] +\Delta \mathbf{Y}_{\text{pred}}$
\end{algorithmic}
\caption{OVECM: Online VECM}
\label{alg:52proposal}
\end{algorithm}

\begin{itemize}
\item The function \texttt{getJohansen} returns cointegration vectors given by
the Johansen method considering the trace statistic test at 95\%
significance level.
\item The function \texttt{vecMatrix} returns the matrices~(\ref{eq:vecA})
and~(\ref{eq:vecY}) that allows VECM to be solved.
\item The function \texttt{vecMatrixOnline} returns the
matrices~(\ref{eq:vecA}) and~(\ref{eq:vecY}) aggregating new data and removing
the old one, avoiding calculation of the matrix $\mathbf{A}$ from scratch.
\item Out-of-sample outputs are saved in the variables 
$\mathbf{Y}_{\text{true}}$ and $\mathbf{Y}_{\text{pred}}$.
\item The model is solved using OLS.
\item In-sample outputs are saved in the variables $\Delta
\mathbf{y}_{\text{true}}$ and $\Delta \mathbf{y}_{\text{pred}}$.
\item The function \texttt{mape} gets the in-sample MAPE for the $l$ time
series.
\item Cointegration vectors are obtained at the beginning and when they are required to be updated. This updating is decided based on the in-sample MAPE of the last $n$ inputs. The average of all
MAPEs are stored in the variable $e$. If the average of MAPEs
($\texttt{mean}(e)$) is above a certain error given by the mean\_error threshold, cointegration vectors are updated.
\item If new cointegration vectors are required, the function
\texttt{vecMatrixUpdate} only updates the corresponding columns of matrix
$\mathbf{A}$ affected by those vectors (see equation~\ref{eq:vecA}).
\end{itemize}
\newpage

A pseudocode of the algorithm \ref{alg:52proposal} is detailed in the figure\ref{fig:pseudocodigo}.
\newline

@@

\section{Results} \label{sec:52results}

@@

\begin{table}[h!]
\caption{Unit roots tests}
\label{tab:52adf}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
& \textbf{Statistic} & \textbf{Critical value} & \textbf{Result}\\
\hline
EURUSD          & -0.64 & -1.94 & True       \\
$\Delta$EURUSD & -70.45   & -1.94 & False       \\
GBPUSD          & -0.63   & -1.94 & True          \\
$\Delta$GBPUSD & -54.53   & -1.94 & False       \\
CHFUSD          & -0.88   & -1.94 & True         \\
$\Delta$CHFUSD & -98.98   & -1.94 & False       \\
JPYUSD          & -0.65 & -1.94 & True        \\
$\Delta$JPYUSD & -85.78 & -1.94 & False     \\ 
\hline
\end{tabular}
\end{center}
\end{table}

Table~\ref{tab:52adf} shows that all currency rates cannot reject the unit root
test but they rejected it with their first differences. This means that all of
them are I(1) time series and we are allowed to use VECM and therefore OVECM.

@@

ARIMA parameters were also obtained using AIC.
Parameters optimisation is presented in table~\ref{tab:52params}:

\begin{table}[ht]
\caption{Parameters optimisation. VECM order and ARIMA parameters were selected
using AIC.}
\label{tab:52params}
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Windows size $L$ & VECM & ARIMA\\
 $L$ & order $(p)$ & order $(p,d,q)$ \\
\hline
100 & 2 & p=2,d=1,q=1\\
400 & 5 & p=1,d=1,q=1\\
700 & 3 &p=2,d=1,q=1\\
1000 &3 & p=2,d=1,q=1\\
\hline
\end{tabular}
\end{center}
\end{table}

@@

Table~\ref{tab:stats} shows in-sample and out-of-sample performance measures:
MAPE, MAE and RMSE for OVECM, SLVECM and SLARIMA. Test were done using the
parameters defined in table \ref{tab:52params}.  We can see that OVECM has very
similar performance than SLVECM and this support the theory that cointegration
vectors vary little in time. Moreover, OVECM also out performed SLARIMA based on
these three performance measures.

@@

\section{Conclusions} \label{sec:52conclusions}
